Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene

### 2023-02-04 manuscript/manuscript.Rmd
rewriting introduction

# Introduction
Existing palaeoclimate reconstructions are generally based on one of three approaches: 1) Quantitative, geochemical proxy data, 2) palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) earth system modelling [@Burgener2023], with the latter ultimately relying on constraints provided by the geological and fossil record [@Lunt2021]. The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022]. To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations. To address these problems, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.

For the marine realm, Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009].

>>>>

# Introduction
The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022]. To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations. To address these challenges, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.

Existing palaeoclimate reconstructions are generally based on one of three approaches. Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca), and sometimes biological proxies such as leaf physiognomy, are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g. @Liu2009; @Royer2012]. These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...]. Palaeoecological and geological proxy data can provide broad climate constraints, but no established transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Kvacek2007, @Cao2019]. Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

Marine, proxy-based, deep-time palaeoclimate reconstructions commonly rely solely on geochemical proxies [@Veizer2015;@Grossman2022]. Despite the impressive size of these archives large spatial gaps bias global temperature reconstructions based on those records [@Jones2022]. In contrast, studies on the terrestrial and coastal climate record have long used ecological and lithological data for quantitative climate reconstructions [@]. For example, a plethora of "nearest living relative" methods exists for deriving numerical climate estmates from taxonomic plant data for the Cenozoic  [@Faucette2007; @...]. These methods work on the the premise that the climatic requirements of ancient plant taxa were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017].

Even when using the full range of available records, any proxy-based climate reconstructions that goes back beyond the Holocene is inevitably very patchy. To infer regional or global-scale patterns from an incomplete record requires the bridging of spatial gaps. This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled.

Here, we apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.

Combining quantitative proxy data with not fully quantified climate constraints, e.g. the presence of tropical weathering products such as bauxite and laterite, has the potential to improve upon existing, purely proxy-based reconstructions [@Scotese2021; @Burgener2023]. Terrestrial and

approach is still rarely employed in deep-time palaeoclimatic studies, as it This requires transforming ordinal or interval-bound data into numerical estimates, which is common

Integrating this type of information into fully quantitative climate reconstructions requires

data provide a different method of reconstructing palaeotemperatures
Palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) 
[@Burgener2023]
For the marine realm, Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009].
